The unphysicality of Hilbert spaces

### 2024-11-26 Articles/2023-HilbertQuantum/Submissions/Quantum Studies/HilbertQuantum.tex
Remove awkward sentence about von Neumann
Final revision after discussion with Francisco; resubmission version

One problem is that most physicists see mathematics as just a tool for calculation, much like oscilloscopes or telescopes are tools to gather data. As long as they do what they are supposed to do, the details are irrelevant because the physics lies somewhere else. If they do not work correctly, it is just a detail for the expert to fix. But mathematics plays a much more fundamental role in physics: it is the language in which we specify our models. When we say, ``States are rays in a Hilbert space,'' we are giving a physical specification. We are saying that there are objects in the physical world that can be modeled with a particular mathematical structure. If Hilbert spaces do not work to model physical states, it is not simply a math problem: the model is wrong, which is a physics problem. This is the question we are examining in this paper: does the math we use in quantum mechanics model the physical world correctly? Is our physical problem well specified?

@@

To summarize, we are not interested here in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. Moreover, the fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, which are physical entities we prepare in a lab. States in physics are the output of a preparation procedure, a characterization that works for both ``pure states'' of mechanics as well as ensembles in statistical mechanics. They are what we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?  
%We note that von Neumann himself placed very high value on conceptual clarity in physical theories.\cite{Redei2006}

>>>>

One problem is that most physicists see mathematics as just a tool for calculation, much like oscilloscopes or telescopes are tools to gather data. As long as they do what they are supposed to do, the details are irrelevant because the physics lies somewhere else. If they do not work correctly, it is just a detail for the expert to fix. But mathematics plays a much more fundamental role in physics: it is the language in which we specify our models. When we say, ``States are rays in a Hilbert space,'' we are providing a specification for a physical system.  We are saying that there is a particular physical system which under certain circumstances can be mathematically modeled in a particular way.  So, if Hilbert spaces do not work, it is not simply a math problem: the model is wrong, which is a physics problem.  This is the question we are examining in this paper: can the math we use in quantum mechanics model a particular class of physical systems correctly?  Is our physical problem well specified?

@@

To summarize, we are not interested here in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. Moreover, the fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, which are physical entities we prepare in a lab. States in physics are the output of a preparation procedure, a characterization that works for both ``pure states'' of mechanics as well as ensembles in statistical mechanics. They are what we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?
